Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-05-19 text/discussion-KS-copy.qmd
discussion revisions

\newpage
\pagenumbering{gobble}
\pagenumbering{arabic}

# Introduction

On January 9, 2019 a public research university purchased College Board *Student Search Service* order no. $448922$, thereby obtaining the contact information of out-of-state prospective students, who would subsequently receive brochures, postcards, email, and targeted social media. This was a large student list purchase, yielding $122,426$ "names." At $\$0.45$ per name, the order cost $\$55,091.70$. The order specified three "search filters" to control which names were purchased: first, the 2020 high school graduating class; second, under the heading "College Board Exams," PSAT score range of 1070 - 1180; and third, under the heading "Geography", the order filtered on 67 "Geomarkets," identified by titles like "IL08 - Northwest Suburbs" and "IL09 - North Shore."^[On the same day, the university purchased two additional orders that filtered on PSAT score ranges of $1190 - 1260$ and $1270 - 1520$, respectively, but selected the same HS graduating class and Geomarkets.] The Geomarket filter carves states and metroplitan areas into smaller geographic units. Upon closer inspection, this student list purchase selected all Geomarkets in the Chicago-land area except for "IL10 - City of Chicago," which is significantly poorer and has a higher share of Black residents. Similarly, in the DC/Maryland/Virginia metro area, the order selected the affluent "MD02 - Montgomery Metropolitan," "MD03 - Central Maryland (w/o Baltimore)," "VA01 - Arlington & Alexandria," and "VA02 - Fairfax County" Geomarkets, but excluded the "MD05 - Prince Georges Metropolitan," "MD07 - Baltimore (Urban)," and "DC01 - District of Columbia" Geomarkets.

@@

Geomarkets carve states and metroplitan areas into smaller geographic units. Upon closer inspection, the three orders selected on all Geomarkets in the Chicago-land area except for "IL10 - City of Chicago", which is significantly poorer and less White. In the DC/Maryland/Virginia metro area, the orders selected the affluent "MD02 - Montgomery Metropolitan," "MD03 - Central Maryland (w/o Baltimore)," "VA01 - Arlington & Alexandria," and "VA02 - Fairfax County" Geomarkets, but excluded the "MD05 - Prince Georges Metropolitan," "MD07 - Baltimore (Urban)," and "DC01 - District of Columbia" Geomarkets, which are poorer and have a higher share of Black residents. An analysis of 500 \[check\] student lists purchased by 9 public research universities found that 11% of purchases filtered on Geomarket [@RN5012]. Recent research suggests that student lists are surprisingly important for college access outcomes and college graduation outcomes, particularly for first-generation and underrepresented minority students [@RN4739; @RN4894]. To the extent that Geomarket borders are correlated with race and class, the Geomarket selections observed here raise concern that Geomarkets -- and other geographic search filters -- may systematically filter working class and non-white students from educational opportunities.
-->
```

Sociology offers several perspectives to describe the process of sorting students into colleges as a process that tends toward social reproduction [@RN4873; @RN2377]. On the student demand side, the status attainment model argues that student postsecondary destination is a function of family socioeconomic background, particularly parental education [e.g., @RN915; @RN1273; @RN1268; @RN5033; @RN1300]. The cultural capital model argues that bourgeois households bestow the pedigree, social networks, and information necessary to claim spots at selective colleges [e.g., @RN1228; @RN1241; @RN4991]. <!--@RN4873 describes schools as "sorting machines" that create, incorporate, and assign students to categories that determine educational opportunities, often reinforcing reinforcing racial and class inequality. -->On the supply-side, the credentialism literature recognizes that colleges have a financial incentive for educational credentials to be required to claim advantageous social and economic positions [@RN1297], and this race for credentials benefits affluent households [e.g., @RN865; @RN861]. Scholarship on enrollment management offers an agentic depiction of college behavior, showing that colleges expend considerable resources recruiting desirable prospects [@RN3519; @RN4758; @cottom2017lower; @RN4324].

@@

The following section introduces background and scholarship about enrollment management and introduces salient concepts from scholarship on quantification. Second, we introduce the case, describing how @RN4982 developed Geomarkets and the Market Segment Model and how these concepts were incorporated into College Board recruiting products. Third, we describe data and methods. Fourth, we present results. \[We find that\ldots ONE OR TWO SENTENCES\]. Finally, we discuss implications for scholarship. Third-party vendors have structured college access for decades and their influence is growing. The contemporary enrollment management industry is characterized by private equity owned firms selling software-as-service platforms that utilize the same logic as @RN4982, for example EAB's "pipeline analytics" machine learning product [vista_eab_ml]. We propose two fruitful approaches to studying the role of third-party vendors in enrollment management. First, case studies -- like this manuscript and @RN5039 -- utilize primary data collection to conduct granular analyses of particular products or vendors. Second, a growing number of subscription databases (e.g., Orbis, Pitchbook, LSEG) serve the information needs of the investment community by providing data on private firms and acquisitions or investments in private firms. These data sources may provide the basis for a scalable "knowledge infrastructure" [@RN5025] that investigates the role of third-party vendors in education.

```{=html}
\[ADD TEXT SOMEWHERE AROUND HERE ABOUT WHY THIS TOPIC IS SIGNIFICANT AND IMPORTANT FOR CURRENT/FUTURE POLICY DEBATES: Why market segment model important. Software as service products dominate em industry and structure college access. Look at eab machine learning product But rarely do we see the recipe underneath these products. The book structuring college access is that recipe for EPS software\]
<!-- 
THIS TEXT IS FINE, BUT TRYING TO SET UP THE CONTRIBUTION TO BROADER QUANTIFICATION LITERATURE, VIS-A-VIS NO-ONE HAS STUDIED MARKET RESEARCH; DON'T THINK THIS IS NECESSARY AND YOU ARE NOT ON FIRM GROUND HERE IN CALLING THIS "MARKET RESEARCH"
Scholarship on quantification in education has focused on consumer-facing measures such as U.S. News Rankings [@RN5003] and UK "league tables" about school quality [@RN5005]. To our knowledge, prior knowledge has not examined business-facing measures produced by third-party market research. Here, the desirable attributes of students and places they live are quantified with the purpose of informing colleges about where they should allocate recruiting resources to efficiently enroll the most desirable students. We reason that market research reproduces inequality by encouraging colleges to plan their recruiting efforts based on a snapshot of social stratification in student demand that is substantially a function of historic, structural inequality in educational opportunity. In contrast to @RN5005, we cannot demonstrate the effect of geomarkets on inequality. However, we show the mechanisms of how geomarkets reproduce inequality by simulating who is included/excluded when geomarkets are utilized within student list purchases as recommended by @RN4982.
-->
```

# Enrollment Management and Quantification

@@

Drawing from scholarship on organizational theory, EM processes require colleges to make many "make or buy" [@RN4738; @RN980] decisions about whether to perform a given task in-house or outsource it to a third-party vendor [@list_biz]. The most widely-known class of EM vendor is EM consulting firms, which provide advice and implementation in the areas of marketing, recruiting, pricing and financial aid, and student success [@RN5027]. As "creating a class" has becomes complicated and high-stakes [@RN3519], many colleges hire EM consulting firms to develop and/or implement recruiting campaigns. The two largest firms -- Ruffalo Noel Levitz and EAB -- claim to serve more than 3,000 colleges and universities collectively [@rnlabout; @eabcomm].^[In our data collection, attempting to obtain data about student list purchases from all public universities in four states, at least half of these universities outsourced student list purchases to an EM consultancy [@list_biz; @list_empirics].]

@@

[^1]: @RN3519 describes the archives of annual reports of admissions activities from a selective liberal arts college. In the 1942 report, "Section 2.5 is titled 'Secondary School Quartile Standings," a measure of where matriculants fell in the rankings of their high school classes" [@RN3519] while Section 3.11 describes the high schools visited. In the chapter entitled "Travel," @RN3519 [p. 69] describes how The College used record-keeping to inform decisions about visits to local high schools in the 2000s: "The College keeps files on hundreds of schools across the country, and around the globe, with which it had contact \ldots The files enable officers to quickly discern"how we've done here lately," \ldots with the number, relative quality, and yield rate of applicants from a particular school."
-->
```

**Quantification**

@RN5007[p. 402] define quantification as "the production and communication of numbers." <!-- The effects of quantification are described by the three themes of reactivity (people and orginizations change behavior in response to salient quantitative measures), discipline (quantification disciplines actors to react in particular ways), and authority (quantification often weakens discretion of local decision-makers).--> The effects of quantification are described by the concepts reactivity, discipline, and authority [@RN5007]. Sociological studies on quantification have predominantly explored how numbers are produced, used, and interpreted across institutional contexts such as administration, democratic rule, economics, personal life and education, with a particular emphasis on rankings [@RN5008; @RN5003; @RN5005]. In the field of critical data studies, quantification is central to scholarship that explores how computation and algorithms structure social opportunities [e.g., @RN4772; @RN4849]. Our study is grounded in the systematic investigation of "interactions between different quantification regimes" across sociology and critical data studies, as called for by @RN5008[p. 239], to explore the role of third-party servicers and products in the sorting of students into colleges. In particular, the discussions of correlation and homophily by @RN4975 show how social reproduction is encoded in algorithms that direct future resources. We explore how correlation and homophily were incorporated into the construction and application of Geomarkets within the Market Segment Model [@RN4982].

```{=html}
<!--Reactivity, discipline, and authority are three interrelated themes that describe the effects of quantification. Reactivity is the idea that salient quantitative measures cause people and organizations to change their behavior. Quantification disciplines actors to react in particular ways. Quantification changes decision-making power authority, often weakening the discretion of local decision-makers. For instance, scholarship about U.S. News & World Report (USNWR) law school rankings demonstrate the effects of quantification [@RN5003; @RN2356; @RN2376; @RN2375]. Law school rankings were developed as a means of informing prospective students about the relative quality of different programs. Rankings affect (reactivity) where students apply and enroll [@RN5003]. Rankings also affected hiring decisions of firms because clients evaluated firms based on the prestige of law schools attended by firm lawyers. Once law schools realized that "important groups of constituents --- students, faculty, trustees, employers, other media --- used rankings to make decisions that had large consequences for schools \ldots [then] schools felt pressured to take them seriously" [@RN5007, p. 415]. Rankings disciplined the behavior of law schools. For example, characteristics valued by the rankings system (e.g., LSAT scores) became more important for decisions about admissions and merit aid [@RN5003]. Rankings also weakened the authority of admissions personnel to make independent decisions about class size and which applicants to admit [@RN5003]. <!--
In a process akin to sports analytics in Moneyball \[CITE MICHAEL LEWIS\], we suggest that market research replaces the wisdom and authority of local admissions counselors.    -->
```

```{=html}
<!--The analysis of UK school "league tables" by @RN5005 shows how quantification can be a mechanism of social reproduction. In 1992, the UK government began publishing school league tables, which ranked schools based on student performance on national exams taken at age 16. League tables facilitate making evaluative comparisons between schools regardless of geographic proximity. They also discipline households and schools to conceive of school quality in terms of test scores, which are substantially a function of the class composition of schools. Using Census data measured at the local authority level (similar to a U.S. county), @RN5005 find that localities with higher performing schools experienced growth in the share of managerial/professional households following the adoption of league tables and a decline in the share of working-class households. Using longitudinal survey data, managerial/professional households were more likely to move to localities with higher ranked schools following the introduction of school league tables. The introduction of league tables contributed to social reproduction because professional/managerial households were more aware of these consumer-facing metrics and had resources to respond by moving to more expensive neighbhorhoods, near higher performing schools. THIS IS RATCHET EFFECT STUFF. I CUT IT BECAUSE YOU HAVEN'T INTRODUCED THE IDEA OF RATCHET EFFECT
Schools that performed well on initial league tables tended to be schools with a high share of professional/managerial families. Therefore, quantification takes a snapshot of existing social stratification and that snapshot becomes a power unto itself, amplifying the effect of previous strctural inequality on subsequent structural inequality.
-->
```

```{=html}
<!-- THIS IS THE CUT TRANSITION PARAGRAPH
In the broader quantification literature, critical data studies [e.g., @RN4975; @RN4849; @RN4772] and sociological analyses of technology and inequality [e.g., @RN4778; @RN4786; @RN4877; @RN4810] introduce concepts salient to the analysis of market research. Our focus is on "selection devices," which utilize some combination of inputs and discretion to categorize places, people, and organizations for the purpose of allocating opportunities or interventions [@RN4778; @RN5012].
-->
```

@@

Due to data limitations, predictive analytics often utilize "proxy" variables -- variables highly correlated with a variable of interest [@RN4849; @RN4975]. For example, "e-scores" utilize proxy variables to identify "people like you" and then predict your buying behavior based on the past buying behavior of people like you. @RN4849 [p. 146] states that "the modelers for e-scores have to make do with trying to answer the question 'How have people like you behaved in the past?' when ideally they would ask, 'How have you behaved in the past?'" We observe similar behavior in market research about college access [e.g., @RN4565], when enrollment managers use 'which colleges did students near you consider' as a proxy for, 'which colleges would you consider?'

Many studies show that predictions based on correlations reproduce structural inequality [for a review see @RN4794]. The correlations observed during the training data stage are a snapshot of relationships between variables at a particular point in time. Observed correlations may be a function of enduring structural inequality, but underlying causes are not considered by applications of predictive models. Reviewing scholarship about algorithms, @RN4794 [p. 224] state that "predicting the future on the basis of the past threatens to reify and reproduce existing inequalities." Disproportionately targeted/excluded populations are predicted to have a higher risk of an outcome, which amplifies subsequent targeting/exclusion. This phenomenon has been termed the "ratchet effect" [@RN4851] and "pernicious feedback loops" [@RN4849].^[An often cited example is the LSI-R recidivism model which predicts a prisoner's chances of re-arrest and is used by 24 states [@RN4849]. Because the algorithm uses zip code as an input, people who live in highly policed neighborhoods have a higher predicted probability of being arrested, which leads to more policing in those neighborhoods, which perpetuates racialized inequality in arrests.] As we discuss below, @RN4982 analyzed 1980 SAT score-sending behavior (training data) and inferred that demand for college is a function of class. They then recommended that selective colleges recruit from localities with many affluent, college-educated households.

@@

Homophily is central to market research products that categorize customers. Geodemography emerged in the 1970s as a branch of market research that estimates the behavior of consumers based on where they live [@RN4878]. Market segments are subgroups within a larger market that have similar consumer demand. Early geodemographic classifications of consumers (e.g., PRIZM by Claritas Corporation) were derived from publicly available Census data, which disaggregated data to the zip code level. The Claritas Corporation had a financial incentive to argue that people living near one another share similar consumer preferences because geographic localities could then be categorized into market segments that would be useful for direct mail marketing campaigns [@RN4957]. Later, the development of individual credit scores (e.g., FICO score) enabled merchants to classify consumers into many, fine-grained groups [@RN4877]. @RN4810 introduce the concept "classification situations" to describe the expansion of actuarial techniques to categorize customers into many, ordinally ranked groups. Merchants and lenders began tying these classifications to tiered products that targets different consumer groups with different levels of benefits and costs [@RN5015].^[For example, "payday loans" charge high interest rates to consumer groups that were previously denied credit.] Classification situations engender markets where a vertical hierarchy of products are matched to a vertical hierarchy of consumers. We observe similar processes in our case study. @RN4982 categorizes students into four market segments -- local, in-state, regional, and national -- based on their score sending behavior and then evaluates the attractiveness of each Geomarket based on how many "regional" and "national" students live there. <!--
ORIGINAL TEXT ABOUT GEODEMOGRAPHIC CLASSIFICATION SYSTEMS

@@

@@

<!--  We make third-party EM products the object of empirical analysis, though we cannot claim to assess how the population of colleges utilize these products.
# REMOVED FROM PARAGRAPH ABOVE: First, in the case of college rankings, extant research shows how consumer-facing quantification disciplines the application/enrollment behavior of prospective students, the hiring behavior of employers, and the behavior of colleges that are disciplined to pursue customers with characteristics valued by rankings [@RN5003; @RN2356; @RN2375].  -->

This manuscript analyzes the Market Segment Model [@RN4982], which categorized high school students into vertical market segments -- local, in-state, regional, national -- and simultaneously created local Geomarkets that could be evaluated based on their composition of student market segments. The Market Segment Model and Geomarkets became the basis for the College Board *Enrollment Planning Service* (EPS), which advised colleges which Geomarkets to target. Later, Geomarkets were incorporated into the College Board student list product named Student Search Service. Unlike the analysis of UK school league tables by @RN5005, we cannot show the effect of quantification on social reproduction. However, by demonstrating which students are included versus excluded in student list purchases that filter on particular Geomarkets, our analyses provide novel insight into the underlying mechanism by which quantification can reproduce historical class-based and race-based inequality in the sorting of students into colleges.

@@

The "initial task" was to define geographic boundaries "in a manner consistent with admissions officers' intuitive understanding of student pools" [@RN4982, p. 4]. Quoting an admissions officer, @RN4982[p. 11] write, "'There are only three kinds of college-bound students: those who want to live at home, those who want to live on campus but bring their laundry home, and those who want to go far enough from home that Mom and Dad can't visit without calling first.'" As such, @RN4982[p. 11] created "three types of boundaries -- region, state, and community." The initial regions in the project were New England, Middle States, and the South. Next, they "divided each state into as few as two and as many as thirty community-based enrollment markets or pools, for a total of 143 separate markets" (p. 11). These enrollment markets, later called Geomarkets, were intended to be consistent with the conception of a catchment market from the perspective of admissions counselors. @RN4982 [pp. 11-12] described the creation of Geomarket borders only briefly:

@@

The thesis of this chapter is that student demand for higher education is a function of social origin. For example, the authors state (p. 33) that "these data allow us to say with considerable confidence that local and in-state students are not likely to come from families in which both parents have received college educations" and that "the implication is simply that college-educated parents instill in their children more wide-ranging educational aspirations." Commenting on family income, @RN4982[p. 33] write that "we could predict that all local students would come from moderate-income or low-income families and be wrong only 5.5 percent of the time." They conclude (p. 42) that "our research has simply demonstrated what everyone has always known: communities with high levels of family income and parental education are also communities in which students have higher than average SATs and more far-reaching aspirations."

>>>>

```{=tex}
\newpage
\pagenumbering{gobble}
\pagenumbering{arabic}
```
# Introduction

On January 9, 2019 a public research university purchased College Board *Student Search Service* order no. $448922$, thereby obtaining the contact information of out-of-state prospective students, who would subsequently receive brochures, postcards, email, and targeted social media. This was a large student list purchase, yielding $122,426$ "names." At $\$0.45$ per name, the order cost $\$55,091.70$. The order specified three "search filters" to control which names were purchased: first, the 2020 high school graduating class; second, under the heading "College Board Exams," PSAT score range of 1070 - 1180; and third, under the heading "Geography", the order filtered on 67 "Geomarkets," identified by titles like "IL08 - Northwest Suburbs" and "IL09 - North Shore."[^1] The Geomarket filter carves states and metroplitan areas into smaller geographic units. Upon closer inspection, this student list purchase selected all Geomarkets in the Chicago-land area except for "IL10 - City of Chicago," which is significantly poorer and has a higher share of Black residents. Similarly, in the DC/Maryland/Virginia metro area, the order selected the affluent "MD02 - Montgomery Metropolitan," "MD03 - Central Maryland (w/o Baltimore)," "VA01 - Arlington & Alexandria," and "VA02 - Fairfax County" Geomarkets, but excluded the "MD05 - Prince Georges Metropolitan," "MD07 - Baltimore (Urban)," and "DC01 - District of Columbia" Geomarkets.

[^1]: On the same day, the university purchased two additional orders that filtered on PSAT score ranges of $1190 - 1260$ and $1270 - 1520$, respectively, but selected the same HS graduating class and Geomarkets.

@@

Geomarkets carve states and metroplitan areas into smaller geographic units. Upon closer inspection, the three orders selected on all Geomarkets in the Chicago-land area except for "IL10 - City of Chicago", which is significantly poorer and less White. In the DC/Maryland/Virginia metro area, the orders selected the affluent "MD02 - Montgomery Metropolitan," "MD03 - Central Maryland (w/o Baltimore)," "VA01 - Arlington & Alexandria," and "VA02 - Fairfax County" Geomarkets, but excluded the "MD05 - Prince Georges Metropolitan," "MD07 - Baltimore (Urban)," and "DC01 - District of Columbia" Geomarkets, which are poorer and have a higher share of Black residents. An analysis of 500 \[check\] student lists purchased by 9 public research universities found that 11% of purchases filtered on Geomarket [@RN5012]. Recent research suggests that student lists are surprisingly important for college access outcomes and college graduation outcomes, particularly for first-generation and underrepresented minority students [@RN4739; @RN4894]. To the extent that Geomarket borders are correlated with race and class, the Geomarket selections observed here raise concern that Geomarkets -- and other geographic search filters -- may systematically filter working class and non-white students from educational opportunities.
-->
```
Sociology offers several perspectives to describe the process of sorting students into colleges as a process that tends toward social reproduction [@RN4873; @RN2377]. On the student demand side, the status attainment model argues that student postsecondary destination is a function of family socioeconomic background, particularly parental education [e.g., @RN915; @RN1273; @RN1268; @RN5033; @RN1300]. The cultural capital model argues that bourgeois households bestow the pedigree, social networks, and information necessary to claim spots at selective colleges [e.g., @RN1228; @RN1241; @RN4991]. <!--@RN4873 describes schools as "sorting machines" that create, incorporate, and assign students to categories that determine educational opportunities, often reinforcing reinforcing racial and class inequality. -->On the supply-side, the credentialism literature recognizes that colleges have a financial incentive for educational credentials to be required to claim advantageous social and economic positions [@RN1297], and this race for credentials benefits affluent households [e.g., @RN865; @RN861]. Scholarship on enrollment management offers an agentic depiction of college behavior, showing that colleges expend considerable resources recruiting desirable prospects [@RN3519; @RN4758; @cottom2017lower; @RN4324].

@@

The following section introduces background and scholarship about enrollment management and introduces salient concepts from scholarship on quantification. Second, we introduce the case, describing how @RN4982 developed Geomarkets and the Market Segment Model and how these concepts were incorporated into College Board recruiting products. Third, we describe data and methods. Fourth, we present results. [We find that\ldots ONE OR TWO SENTENCES]. Finally, we discuss implications for scholarship. Third-party vendors have structured college access for decades and their influence is growing. The contemporary enrollment management industry is characterized by private equity owned firms selling software-as-service platforms that utilize the same logic as @RN4982, for example EAB's "pipeline analytics" machine learning product [vista_eab_ml]. We propose two fruitful approaches to studying the role of third-party vendors in enrollment management. First, case studies -- like this manuscript and @RN5039 -- utilize primary data collection to conduct granular analyses of particular products or vendors. Second, a growing number of subscription databases (e.g., Orbis, Pitchbook, LSEG) serve the information needs of the investment community by providing data on private firms and acquisitions or investments in private firms. These data sources may provide the basis for a scalable "knowledge infrastructure" [@RN5025] that investigates the role of third-party vendors in education.

```{=html}
\[ADD TEXT SOMEWHERE AROUND HERE ABOUT WHY THIS TOPIC IS SIGNIFICANT AND IMPORTANT FOR CURRENT/FUTURE POLICY DEBATES: Why market segment model important. Software as service products dominate em industry and structure college access. Look at eab machine learning product But rarely do we see the recipe underneath these products. The book structuring college access is that recipe for EPS software\]
<!-- 
THIS TEXT IS FINE, BUT TRYING TO SET UP THE CONTRIBUTION TO BROADER QUANTIFICATION LITERATURE, VIS-A-VIS NO-ONE HAS STUDIED MARKET RESEARCH; DON'T THINK THIS IS NECESSARY AND YOU ARE NOT ON FIRM GROUND HERE IN CALLING THIS "MARKET RESEARCH"
Scholarship on quantification in education has focused on consumer-facing measures such as U.S. News Rankings [@RN5003] and UK "league tables" about school quality [@RN5005]. To our knowledge, prior knowledge has not examined business-facing measures produced by third-party market research. Here, the desirable attributes of students and places they live are quantified with the purpose of informing colleges about where they should allocate recruiting resources to efficiently enroll the most desirable students. We reason that market research reproduces inequality by encouraging colleges to plan their recruiting efforts based on a snapshot of social stratification in student demand that is substantially a function of historic, structural inequality in educational opportunity. In contrast to @RN5005, we cannot demonstrate the effect of geomarkets on inequality. However, we show the mechanisms of how geomarkets reproduce inequality by simulating who is included/excluded when geomarkets are utilized within student list purchases as recommended by @RN4982.
-->
```
# Enrollment Management and Quantification

@@

Drawing from scholarship on organizational theory, EM processes require colleges to make many "make or buy" [@RN4738; @RN980] decisions about whether to perform a given task in-house or outsource it to a third-party vendor [@list_biz]. The most widely-known class of EM vendor is EM consulting firms, which provide advice and implementation in the areas of marketing, recruiting, pricing and financial aid, and student success [@RN5027]. As "creating a class" has becomes complicated and high-stakes [@RN3519], many colleges hire EM consulting firms to develop and/or implement recruiting campaigns. The two largest firms -- Ruffalo Noel Levitz and EAB -- claim to serve more than 3,000 colleges and universities collectively [@rnlabout; @eabcomm].[^2]

[^2]: In our data collection, attempting to obtain data about student list purchases from all public universities in four states, at least half of these universities outsourced student list purchases to an EM consultancy [@list_biz; @list_empirics].

@@

[^1]: @RN3519 describes the archives of annual reports of admissions activities from a selective liberal arts college. In the 1942 report, "Section 2.5 is titled 'Secondary School Quartile Standings," a measure of where matriculants fell in the rankings of their high school classes" [@RN3519] while Section 3.11 describes the high schools visited. In the chapter entitled "Travel," @RN3519 [p. 69] describes how The College used record-keeping to inform decisions about visits to local high schools in the 2000s: "The College keeps files on hundreds of schools across the country, and around the globe, with which it had contact \ldots The files enable officers to quickly discern"how we've done here lately," \ldots with the number, relative quality, and yield rate of applicants from a particular school."
-->
```
**Quantification**

@RN5007 [p. 402] define quantification as "the production and communication of numbers." <!-- The effects of quantification are described by the three themes of reactivity (people and orginizations change behavior in response to salient quantitative measures), discipline (quantification disciplines actors to react in particular ways), and authority (quantification often weakens discretion of local decision-makers).--> The effects of quantification are described by the concepts reactivity, discipline, and authority [@RN5007]. Sociological studies on quantification have predominantly explored how numbers are produced, used, and interpreted across institutional contexts such as administration, democratic rule, economics, personal life and education, with a particular emphasis on rankings [@RN5008; @RN5003; @RN5005]. In the field of critical data studies, quantification is central to scholarship that explores how computation and algorithms structure social opportunities [e.g., @RN4772; @RN4849]. Our study is grounded in the systematic investigation of "interactions between different quantification regimes" across sociology and critical data studies, as called for by @RN5008 \[p. 239\], to explore the role of third-party servicers and products in the sorting of students into colleges. In particular, the discussions of correlation and homophily by @RN4975 show how social reproduction is encoded in algorithms that direct future resources. We explore how correlation and homophily were incorporated into the construction and application of Geomarkets within the Market Segment Model [@RN4982].
<!--Reactivity, discipline, and authority are three interrelated themes that describe the effects of quantification. Reactivity is the idea that salient quantitative measures cause people and organizations to change their behavior. Quantification disciplines actors to react in particular ways. Quantification changes decision-making power authority, often weakening the discretion of local decision-makers. For instance, scholarship about U.S. News & World Report (USNWR) law school rankings demonstrate the effects of quantification [@RN5003; @RN2356; @RN2376; @RN2375]. Law school rankings were developed as a means of informing prospective students about the relative quality of different programs. Rankings affect (reactivity) where students apply and enroll [@RN5003]. Rankings also affected hiring decisions of firms because clients evaluated firms based on the prestige of law schools attended by firm lawyers. Once law schools realized that "important groups of constituents --- students, faculty, trustees, employers, other media --- used rankings to make decisions that had large consequences for schools \ldots [then] schools felt pressured to take them seriously" [@RN5007, p. 415]. Rankings disciplined the behavior of law schools. For example, characteristics valued by the rankings system (e.g., LSAT scores) became more important for decisions about admissions and merit aid [@RN5003]. Rankings also weakened the authority of admissions personnel to make independent decisions about class size and which applicants to admit [@RN5003]. <!--
In a process akin to sports analytics in Moneyball \[CITE MICHAEL LEWIS\], we suggest that market research replaces the wisdom and authority of local admissions counselors.    -->
<!--The analysis of UK school "league tables" by @RN5005 shows how quantification can be a mechanism of social reproduction. In 1992, the UK government began publishing school league tables, which ranked schools based on student performance on national exams taken at age 16. League tables facilitate making evaluative comparisons between schools regardless of geographic proximity. They also discipline households and schools to conceive of school quality in terms of test scores, which are substantially a function of the class composition of schools. Using Census data measured at the local authority level (similar to a U.S. county), @RN5005 find that localities with higher performing schools experienced growth in the share of managerial/professional households following the adoption of league tables and a decline in the share of working-class households. Using longitudinal survey data, managerial/professional households were more likely to move to localities with higher ranked schools following the introduction of school league tables. The introduction of league tables contributed to social reproduction because professional/managerial households were more aware of these consumer-facing metrics and had resources to respond by moving to more expensive neighbhorhoods, near higher performing schools. THIS IS RATCHET EFFECT STUFF. I CUT IT BECAUSE YOU HAVEN'T INTRODUCED THE IDEA OF RATCHET EFFECT
Schools that performed well on initial league tables tended to be schools with a high share of professional/managerial families. Therefore, quantification takes a snapshot of existing social stratification and that snapshot becomes a power unto itself, amplifying the effect of previous strctural inequality on subsequent structural inequality.
<!-- THIS IS THE CUT TRANSITION PARAGRAPH
In the broader quantification literature, critical data studies [e.g., @RN4975; @RN4849; @RN4772] and sociological analyses of technology and inequality [e.g., @RN4778; @RN4786; @RN4877; @RN4810] introduce concepts salient to the analysis of market research. Our focus is on "selection devices," which utilize some combination of inputs and discretion to categorize places, people, and organizations for the purpose of allocating opportunities or interventions [@RN4778; @RN5012].

@@

Due to data limitations, predictive analytics often utilize "proxy" variables -- variables highly correlated with a variable of interest [@RN4849; @RN4975]. For example, "e-scores" utilize proxy variables to identify "people like you" and then predict your buying behavior based on the past buying behavior of people like you. @RN4849 \[p. 146\] states that "the modelers for e-scores have to make do with trying to answer the question 'How have people like you behaved in the past?' when ideally they would ask, 'How have you behaved in the past?'" We observe similar behavior in market research about college access [e.g., @RN4565], when enrollment managers use 'which colleges did students near you consider' as a proxy for, 'which colleges would you consider?'

Many studies show that predictions based on correlations reproduce structural inequality [for a review see @RN4794]. The correlations observed during the training data stage are a snapshot of relationships between variables at a particular point in time. Observed correlations may be a function of enduring structural inequality, but underlying causes are not considered by applications of predictive models. Reviewing scholarship about algorithms, @RN4794 [p. 224] state that "predicting the future on the basis of the past threatens to reify and reproduce existing inequalities." Disproportionately targeted/excluded populations are predicted to have a higher risk of an outcome, which amplifies subsequent targeting/exclusion. This phenomenon has been termed the "ratchet effect" [@RN4851] and "pernicious feedback loops" [@RN4849].[^3] As we discuss below, @RN4982 analyzed 1980 SAT score-sending behavior (training data) and inferred that demand for college is a function of class. They then recommended that selective colleges recruit from localities with many affluent, college-educated households.

[^3]: An often cited example is the LSI-R recidivism model which predicts a prisoner's chances of re-arrest and is used by 24 states [@RN4849]. Because the algorithm uses zip code as an input, people who live in highly policed neighborhoods have a higher predicted probability of being arrested, which leads to more policing in those neighborhoods, which perpetuates racialized inequality in arrests.

@@

Homophily is central to market research products that categorize customers. Geodemography emerged in the 1970s as a branch of market research that estimates the behavior of consumers based on where they live [@RN4878]. Market segments are subgroups within a larger market that have similar consumer demand. Early geodemographic classifications of consumers (e.g., PRIZM by Claritas Corporation) were derived from publicly available Census data, which disaggregated data to the zip code level. The Claritas Corporation had a financial incentive to argue that people living near one another share similar consumer preferences because geographic localities could then be categorized into market segments that would be useful for direct mail marketing campaigns [@RN4957]. Later, the development of individual credit scores (e.g., FICO score) enabled merchants to classify consumers into many, fine-grained groups [@RN4877]. @RN4810 introduce the concept "classification situations" to describe the expansion of actuarial techniques to categorize customers into many, ordinally ranked groups. Merchants and lenders began tying these classifications to tiered products that targets different consumer groups with different levels of benefits and costs [@RN5015].[^4] Classification situations engender markets where a vertical hierarchy of products are matched to a vertical hierarchy of consumers. We observe similar processes in our case study. @RN4982 categorizes students into four market segments -- local, in-state, regional, and national -- based on their score sending behavior and then evaluates the attractiveness of each Geomarket based on how many "regional" and "national" students live there. <!--
ORIGINAL TEXT ABOUT GEODEMOGRAPHIC CLASSIFICATION SYSTEMS

@@

[^4]: For example, "payday loans" charge high interest rates to consumer groups that were previously denied credit.

@@

```{=html}
<!--  We make third-party EM products the object of empirical analysis, though we cannot claim to assess how the population of colleges utilize these products.
# REMOVED FROM PARAGRAPH ABOVE: First, in the case of college rankings, extant research shows how consumer-facing quantification disciplines the application/enrollment behavior of prospective students, the hiring behavior of employers, and the behavior of colleges that are disciplined to pursue customers with characteristics valued by rankings [@RN5003; @RN2356; @RN2375].  -->
```
This manuscript analyzes the Market Segment Model [@RN4982], which categorized high school students into vertical market segments -- local, in-state, regional, national -- and simultaneously created local Geomarkets that could be evaluated based on their composition of student market segments. The Market Segment Model and Geomarkets became the basis for the College Board *Enrollment Planning Service* (EPS), which advised colleges which Geomarkets to target. Later, Geomarkets were incorporated into the College Board student list product named Student Search Service. Unlike the analysis of UK school league tables by @RN5005, we cannot show the effect of quantification on social reproduction. However, by demonstrating which students are included versus excluded in student list purchases that filter on particular Geomarkets, our analyses provide novel insight into the underlying mechanism by which quantification can reproduce historical class-based and race-based inequality in the sorting of students into colleges.

@@

The "initial task" was to define geographic boundaries "in a manner consistent with admissions officers' intuitive understanding of student pools" [@RN4982, p. 4]. Quoting an admissions officer, @RN4982 \[p. 11\] write, "'There are only three kinds of college-bound students: those who want to live at home, those who want to live on campus but bring their laundry home, and those who want to go far enough from home that Mom and Dad can't visit without calling first.'" As such, @RN4982 [p. 11] created "three types of boundaries -- region, state, and community." The initial regions in the project were New England, Middle States, and the South. Next, they "divided each state into as few as two and as many as thirty community-based enrollment markets or pools, for a total of 143 separate markets" (p. 11). These enrollment markets, later called Geomarkets, were intended to be consistent with the conception of a catchment market from the perspective of admissions counselors. @RN4982 [pp. 11-12] described the creation of Geomarket borders only briefly:

@@

The thesis of this chapter is that student demand for higher education is a function of social origin. For example, the authors state (p. 33) that "these data allow us to say with considerable confidence that local and in-state students are not likely to come from families in which both parents have received college educations" and that "the implication is simply that college-educated parents instill in their children more wide-ranging educational aspirations." Commenting on family income, @RN4982 [p. 33] write that "we could predict that all local students would come from moderate-income or low-income families and be wrong only 5.5 percent of the time." They conclude (p. 42) that "our research has simply demonstrated what everyone has always known: communities with high levels of family income and parental education are also communities in which students have higher than average SATs and more far-reaching aspirations."

The authors also find that these four variables predict student score-sending behavior at the Geomarket-level. However, Geomarkets differ in the relative abundance of students with particular socioeconomic characteristics, which has practical implications for recruiting. Therefore, @RN4982 [p. 44] recommend that colleges target Geomarkets with desirable compositions of socioeconomic characteristics in order to reach students from desired student market segments:
